Stability and Change in Personal Culture

### 2019-12-09 manuscript9.Rmd

---
title: "Stability and Change in Personal Culture"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

@@

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).

In this paper, we make these two models of attitude change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data. In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). By classifying the pattern of change at the individual level, we can clarify when different accounts of aggregate change are more likely to apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at bars or attending church, that are important in contemporary U.S. society.

@@

Equation \@ref(eq:AR) represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time. Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state. This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin].

@@

The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007]. The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level. These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions (Mannheim 1952; Ryder 1965). Vaisey and Lizardo (2016), looking at a range of opinion items in the GSS, suggest that change for most likely comes in the form of cohort replacement.

@@

Another possibility is that participation in the survey itself produces change or stability, referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal]. While Warren and Halpern-Manners [@-warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns. One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster. If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 becomes a better predictor of wave 3.

@@

Perhaps the most striking pattern in our findings pertains to questions about _gay rights_. There are 6 items that ask about some aspect of gay rights, and all show evidence for active updating. Questions about civil liberties for gays and about gay marriage are the highest in their categories. The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views. This pattern is strongly consistent with Zaller's [@-zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.

@@

First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model. Around 40 percent of all items show no evidence for updating. For the items that do show evidence of active updating, the overall rate of change in the population is likely low. Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to where they were. What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago. Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is. But the average level of updating for most views appears to be low.

@@

In contrast to other beliefs, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging people to update their views. By definition, salience is a limited resource, meaning only a few beliefs and behaviors could change at this rate during any given period. The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views. This general model is more consistent with a Bourdieusian theory that emphasizes the "conditions of past production" rather than processes of active meaning construction with little long-term memory.

@@

For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change. Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors and institutional norms that encourage openness to change at different ages. Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes. Doing so will expand our understanding of the institutional and developmental factors that that give rise to stable or variable attitudes [@howekrosnick]. Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is an important step in that direction [@eatonetal; @vissermirabile].

@@

Our work cannot speak to the strength of ideological commitments or whether individuals have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on specific policy questions. The low degree of consistency in beliefs is consistent with the idea that people are "ideologically innocent" [@kinderkalmoe]. What we can say is that individuals over 30 in American during the period we studied did not make lasting changes in whether and how strongly they thought of themselves as "liberal" or "conservative."

That said, respondents are not wholly without consistent opinions. On items such spending on space and welfare, divorce laws, and the death penalty, individuals' responses are consistent from one wave to the next. While some of these items, particularly those pertaining to gay marriage and public spending on welfare, received significant media attention during the period covered by the panel, and thus are consistent with Zaller's [@zaller] model of attitude formation, space exploration was not a high-profile issue in this time period, and we struggle to account for why this item exhibits a comparatively high $\beta$ estimate.

>>>>

---
title: "Stability and Change in Personal Culture"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs and behaviors in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

@@

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). These two models represent different approaches to understanding how people come to hold different forms of _personal culture_, or the declarative and non-delcarative attitudes, worldviews, values, dispositions, and associations that manifest at the individual level [@lizardo2016].

In this paper, we make these two models of personal culture change explicit, deduce some of their empirical implications, and derive a statistical model for estimating the prevalence of active updating using panel data. In contrast to previous models that tend to assume one of these data-generating processes to measure stability and reliability of estimates over time, our approach separates persisting change from non-persisting change to estimate if there is evidence of individuals in the population making persistent changes in attitudes and behaviors. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). By classifying the pattern of change in personal culture, we can clarify when different accounts of aggregate change are more likely to apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at bars or attending church, that are important in contemporary U.S. society.

@@

Equation \@ref(eq:AR) represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time. Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state. This formal property is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin]. In this framework, we use the term "updating" to refer to the change to a new baseline for whatever reason, not necessarily requiring the incorporation of new or better information.

@@

The SDM underlies theories that suggest that individuals are stable in their dispositions throughout life or, if they do change, tend to revert back to a relatively stable baseline in short time, including “control” theories such as Identity Control Theory and Affect Control Theory [@robinson2007; @smithlovinheise]. The notion that individuals are stable in their dispositions is commonly associated with cohort replacement theories of attitude change at the population level. These models posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions (Mannheim 1952; Ryder 1965). Vaisey and Lizardo (2016), looking at a range of opinion items in the GSS, suggest that change for most likely comes in the form of cohort replacement.

@@

Another possibility is that participation in the survey itself produces change or stability, referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal]. While Warren and Halpern-Manners [-@warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns. One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystallize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to “game” the survey to get through it faster. If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as respondents would change between waves 1 and 2, and wave 2 becomes a better predictor of wave 3.

@@

Perhaps the most striking pattern in our findings pertains to questions about _gay rights_. There are 6 items that ask about some aspect of gay rights, and all show evidence for active updating. Questions about civil liberties for gays and about gay marriage are the highest in their categories. The huge public and political salience of this issue throughout the study period likely made this issue one where more people than usual were open to revising their views. This pattern is strongly consistent with Zaller's [-@zaller] argument that highly salient issues where elite opinion shifts can lead to large changes in public opinion.

@@

First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model. Around 40 percent of all items show no evidence for updating. For the items that do show evidence of active updating, the overall rate of change in the population is likely low. Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to a stable baseline. What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago. Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is. But the average level of updating for most views appears to be low.

@@

The overall pattern of results suggests that one reason attitudes are largely stable is because most issues simply do not reach the kind of salience necessary to shift people's opinions. In contrast to other beliefs, the pattern of findings for gay rights show that a high degree of public salience and social movements can accelerate change by encouraging people to update their views. By definition, salience is a limited resource, meaning only a few beliefs and behaviors could change at this rate during any given period. The baseline process of attitude change appears to be more consistent with a model that shows that people do not really change; rather, they die and are replaced by cohorts with different views. This general model is more consistent with a Bourdieusian theory that emphasizes the "conditions of past production" rather than processes of active meaning construction with little long-term memory.

@@

For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change. This pattern suggests that the salience of an issue can matter at the individual level, as well as the societal level. Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors and institutional norms that encourage openness to change at different ages. Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes. Doing so will expand our understanding of the institutional and developmental factors that that give rise to stable or variable attitudes [@howekrosnick]. Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is an important step in that direction [@eatonetal; @vissermirabile].

@@

The finding that ideological identification is comparatively stable should not be taken to imply that most people in the population have the kind of tightly knit belief structures that political scientists typically call "ideologies." Outside of a handful of high-profile political items such as partisan identity, abortion, and gay marriage, individuals appear to lack strong, clear opinions on most specific policy questions. The low degree of consistency in beliefs is consistent with the idea that people are "ideologically innocent" [@kinderkalmoe]. What we can say is that individuals over 30 in American during the period we studied did not make lasting changes in whether and how strongly they thought of themselves as "liberal" or "conservative."

That said, respondents are not wholly without consistent opinions. On items such spending on space and welfare, divorce laws, and the death penalty, individuals' responses are consistent from one wave to the next. While some of these items, particularly those pertaining to gay marriage and public spending on welfare, received significant media attention during the period covered by the panel, and thus are consistent with Zaller's [-@zaller] model of attitude formation, space exploration was not a high-profile issue in this time period, and we struggle to account for why this item exhibits a comparatively high $\beta$ estimate.
